---
title: "The Price of Weather"
author: "Sampson Wescoatt"
date: "12/17/2020"
output: html_document
---

## Synopsis

***

## Data Processing

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, message = FALSE)
```

Data for this assignment is provided in .bz2 compressed file format, so I used the read.csv() function to extract the data and load it into memory as a tibble using the dplyr library for effective data manipulation.

```{r, cache = TRUE}
## The dplyr library in tidyverse provides simple, effective data processing.
library(dplyr)

## The read.csv() function can take .bz2 files directly.
sdata <- as_tibble(read.csv("./repdata_data_StormData.csv.bz2"))
```

***

## Analysis

** Across the United States, which types of events are the most harmful with respect to population health? **

To evaluate this question, I grouped the raw data by event type and counted the fatalities and injuries by event type.

```{r}
library(tidyr)

fatal_data <- sdata %>% 
  
  ## Select the appropriate metrics
  select(EVTYPE, FATALITIES, INJURIES) %>% 
  
  ## Group by event
  group_by(event = EVTYPE) %>%
  
  ## Produce summary of harm
  summarize(
    fatalities = sum(FATALITIES, na.rm = TRUE),
    injuries = sum(INJURIES, na.rm = TRUE),
    total = fatalities + injuries) %>%
  
  ## Categorize totals by type
  pivot_longer(cols = c("fatalities", "injuries"), 
               names_to = "type", values_to = "count") %>%
  
  ## Sort by most total fatalities + injuries
  arrange(desc(total)) %>%
  
  ## Select the top 15 events (x2 rows per event)
  head(30)
```


** Across the United States, which types of events have the greatest economic consequences? **

To answer this question, I group the raw data by event type once more, and summarized the property and crop damage.

```{r}
econ_dmg <- sdata %>%
  
  ## Select the appropriate metrics
  select(EVTYPE, PROPDMG, PROPDMGEXP, CROPDMG, CROPDMGEXP) %>%
  
  ## Multiply base by exponent
  mutate(prop_total = ifelse(PROPDMGEXP == "B", PROPDMG * 1000000000,
                          ifelse(PROPDMGEXP == "M", PROPDMG * 1000000,
                              ifelse(PROPDMGEXP == "K", PROPDMG * 1000, 
                                     PROPDMG)))) %>%
  
  ## Perform the same conversion for crop damage
  mutate(crop_total = ifelse(CROPDMGEXP == "B", CROPDMG * 1000000000,
                          ifelse(CROPDMGEXP == "M", CROPDMG * 1000000,
                              ifelse(CROPDMGEXP == "K", CROPDMG * 1000, 
                                     CROPDMG)))) %>%
  
  ## Group by event type
  group_by(event = EVTYPE) %>%
  
  ## Summarize by total damage
  summarize(prop_dmg = sum(prop_total, na.rm = TRUE),
            crop_dmg = sum(crop_total, na.rm = TRUE),
            dmg_total = prop_dmg + crop_dmg) %>%
  
  ## Categorize totals by type
  pivot_longer(cols = c("prop_dmg", "crop_dmg"),
               names_to = "type", values_to = "cost") %>%
  
  ## Sort by highest cost
  arrange(desc(dmg_total)) %>%
  
  ## Select the top 15 (x2 rows per event)
  head(30)
```


***

## Results

```{r}
library(ggplot2)

fatal_plot <- ggplot(fatal_data, aes(count, event, fill = type)) + 
  geom_bar(stat = "identity", position = position_dodge()) +
  labs(title = "Effects of Severe Weather on Population Health",
       subtitle = "Top 15 Weather Events by Fatalities & Injuries",
       caption = "Data from the NOAA Storm Database",
       x = "",
       y = "")

print(fatal_plot)
```


```{r}
econ_plot <- ggplot(econ_dmg, aes(cost, event, fill = type)) + 
  geom_bar(stat = "identity", position = position_dodge()) +
  labs(title = "Economic Impact of Severe Weather Events",
       subtitle = "Top 15 Weather Events by Property & Crop Damage",
       caption = "Data from the NOAA Storm Database",
       x = "",
       y = "")

print(econ_plot)
```